Expanding the Power of Lens Synthesis \\ (talk proposal)

### 2019-02-12 papers/BXProposal/main.tex
wordsmithing and updated abstract

\begin{abstract}
emph{AFM: I will remove this from the submission, but am including it in the
    document for people to review. It is due today (February 12)}
Recent work has shown how to synthesize a class of string bijections from
regular expression specifications. These bijective string transformations
can be useful for synchronizing simple data formats, but many interesting
formats require non-bijective lenses. In this talk, we show how to expand
the class of synthesized transformations beyond bijections to a large subset
of full symmetric lenses (including all the standard asymmetric lenses)
using lightweight user annotations that help identify a ``bijective core.''
\end{abstract}

@@

Bijections are an important class of bidirectional transformations, but many
useful transformations are not completely bijective. Boomerang expressions are
not necessarily bijections -- they encode lenses~\cite{Focal2005-long2}, and
various lens extensions like quotient lenses~\cite{quotientlenses}. Lenses are
bidirectional transformations that may lose information when going from one side
to the other, while obeying certain roundtripping properties. Quotient lenses
obey similar properties to lenses, but also permit treating certain pieces of
information as irrelevant -- normalizing them to a canonical representative.

While we wish to make creating these lenses easier, Optician's synthesis algorithm
only works on bijective lenses. However, lenses and quotient lenses seem to have
a ``bijective core'', a component that merely changes the shape of data.
This bijective component can then be surrounded by projection and quotient
components to form a full-fledged quotient lens.

At its base, surrounding bijections in this way can be done solely with Optician
and Boomerang, but doing so requires repeated code. For example, consider the
example of converting an index of musical composers (with names, birth and death
dates, and nationalities) to a simple index of musical composers (with just
names and nationalities), but with irrelevant whitespace. In this example, ``.''
represents concatenation and ``\lstinline{*}'' represents the Kleene star.

@@

In this example, the bijective core is the bijection that transmits nationality and
names (while swapping them), the projection component removes \lstinline{dates},
and the quotient component normalizes the whitespace.

@@

We can also build a normalization component for \lstinline{dates} using existing
Boomerang combinators.
%
\begin{lstlisting}
  let ws_norm = normalize (del " "*)
  let nodates_norm = iterate (nationality . (rquot ws_norm) . name)
\end{lstlisting}
%
Here, \lstinline{normalize} turns a lens into a ``canonizer'' (in effect, a
normalization function). In this example, \lstinline{normalize} canonizes an
arbitrarily long list of whitespace to the empty string. Then, the canonizer is
hooked into a lens with \lstinline{rquot}, essentially normalizing strings when
the lens is being operated from right-to-left.

@@

Furthermore, we think this approach could scale to other lens extensions, like
matching lenses and dictionary lenses, and other types of bidirectional
constructs beyond merely lenses. All that is needed for this approach is a
user-provided a reduction to a ``bijective core,'' and a means to synthesize
that core. This means that, if a bijective synthesis algorithm is discovered for
(for example) trees or graphs, then annotations can be used to synthesize more
complex bidirectional programs for those structures as well.

>>>>

\begin{abstract}
  {\color{dkblue} \emph{AFM: I will remove this from the submission, but am including it in the
    document for people to review. It is due today (February 12)}} Recent work
  has shown how to synthesize a class of string bijections from regular
  expression specifications. These bijective string transformations can be
  useful for synchronizing simple data formats, but many interesting formats
  require non-bijective lenses. In this talk, we show how to expand the class of
  synthesized transformations beyond bijections to quotient lenses and a large
  subset of full symmetric lenses (including all the standard asymmetric lenses)
  through lightweight user annotations that help identify a ``bijective core.''
\end{abstract}

@@

Bijections are an important class of bidirectional transformations, but many
useful transformations are not completely bijective. In particular, Boomerang
expressions are not necessarily bijections -- they encode lenses and
various lens extensions like quotient lenses~\cite{quotientlenses}. Lenses are
bidirectional transformations that may lose information when going from one side
to the other, while obeying certain roundtripping properties. Quotient lenses
obey similar properties to lenses, but also permit treating certain pieces of
information as irrelevant -- normalizing them to a canonical representative.

While we wish to make creating these lenses easier, Optician's synthesis
algorithm only works on bijective lenses. Fortunately, many lenses and quotient
lenses seem to have a ``bijective core'', a component that merely changes the
shape of data, but does not remove any data. This bijective component can then
be surrounded by projection and quotient components to form a full-fledged
quotient lens.

Surrounding bijections in this way can be already be done with Optician
and Boomerang, but doing so requires repeated code. Consider the
converting an index of musical composers (with names, birth and death
dates, and nationalities) to a simple index of musical composers (with just
names and nationalities and irrelevant whitespace). In this example, ``.''
represents concatenation and ``\lstinline{*}'' represents the Kleene star.

@@

In this example, the bijective core is the bijection that transmits nationality and
names (while swapping them), the projection component removes \lstinline{dates}
when operating from left-to-right and recovers it when operating from right-to-left,
and the quotient component normalizes the whitespace to the empty string.

@@

We can also build a normalization component for \lstinline{dates} using existing
Boomerang combinators.
%
\begin{lstlisting}
  let ws_norm = normalize (del " "*)
  let nodates_norm = iterate (nationality . (rquot ws_norm) . name)
\end{lstlisting}
%
Here, \lstinline{normalize} turns a lens into a ``canonizer'' (in effect, a
normalization function). In this example, \lstinline{normalize} canonizes an
arbitrarily long list of whitespace to the empty string. Then, the canonizer is
hooked into a lens with \lstinline{rquot}, normalizing strings when the lens is
applied right-to-left, and performing the identity when applied from left-to-right.

@@

Furthermore, we think this approach could scale to other lens extensions, like
matching lenses and dictionary lenses, and other types of bidirectional
constructs beyond merely lenses. All that is needed for this approach is a
user-provided a reduction to a ``bijective core,'' and a means to synthesize
that core. If a bijective synthesis algorithm is discovered for
(for example) trees or graphs, then annotations can be used to synthesize more
complex bidirectional programs for those structures as well.
